---
output: 
  html_document:  
    mathjax: null
    style: "yeti"
    toc: true 
    toc_float: true
params:
  doc_title: "Pavian metagenomics classification report"
  doc_author: !r paste("generated with Pavian v", utils::packageVersion("pavian"))
  doc_date: !r date()
  set_name: "example_data"
  all_data_loaded: FALSE
  sample_data: NA
  reports: NA
  include_sankey: TRUE
  filter_taxa: NA
---

---
title: "`r params$doc_title`"
author: "`r params$doc_author`"
date: "`r params$doc_date`"
---

```{r setup, include=FALSE}
library(pavian)
library(shiny)
library(DT)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(DT.options = list(pageLength = 15,
                          saveState = TRUE,
                          searchHighlight = TRUE,
                          scrollX = TRUE,
                          colReorder = TRUE,
                          selection = 'single',
                          dom = 'Bfrtip',
                          class = "stripe hover row-border compact",
                          lengthMenu = list(c(15, 25, 50, 100), c('15','25','50','100')),
                          search = list(regex = TRUE, caseInsensitive = TRUE)))
set_name <- params$set_name
```

Sample set summary
=====================================  

```{r load_data}
if (isTRUE(params$all_data_loaded)) {
  sample_data <- params$sample_data
  reports     <- params$reports
} else {
  sample_directory <- system.file("shinyapp","example-data","brain-biopsies", package = "pavian")
  sample_data <- pavian::read_sample_data(sample_directory)
  reports     <- pavian::read_reports(sample_data$ReportFilePath, sample_data$Name)
}

samples_summary <- pavian::summarize_reports(reports)
samples_summary$Name <- rownames(samples_summary)
extra_cols <- c("Name")
samples_summary <- samples_summary[, c(extra_cols, setdiff(colnames(samples_summary), extra_cols))]
colnames(samples_summary) <- pavian:::beautify_string(colnames(samples_summary))
```

## Run information

```{r run_info, results='asis'}
total_reads <- tryCatch(sum(samples_summary[, 2], na.rm = TRUE), error = function(e) NA)
cat(sprintf("
| | |
|---|---|
| **Sample set** | %s |
| **Number of samples** | %d |
| **Total reads (across samples)** | %s |
| **Filtered taxa (Sankey)** | %s |
| **Generated** | %s |
| **Pavian version** | %s |
",
  params$set_name,
  length(reports),
  ifelse(is.na(total_reads), "NA", format(total_reads, big.mark = ",")),
  ifelse(length(params$filter_taxa) == 0 || all(is.na(params$filter_taxa)),
         "Chordata, other sequences (default)",
         paste(params$filter_taxa, collapse = ", ")),
  format(Sys.time(), "%Y-%m-%d %H:%M %Z"),
  as.character(utils::packageVersion("pavian"))
))
```

## Tables

```{r sample_summary}
start_color_bar_at <- 3
samples_summary_percent <- samples_summary
samples_summary_percent[, start_color_bar_at:ncol(samples_summary)] <-
  100 * signif(sweep(samples_summary[, start_color_bar_at:ncol(samples_summary)],
                     1, samples_summary[, 2], `/`), 4)

tabsetPanel(
  tabPanel("Classification summary (%)",
    DT::datatable(samples_summary_percent, rownames = FALSE,
                  class = "stripe hover row-border compact",
                  extensions = 'Buttons',
                  options = list(dom = 'Bfrtip',
                                 buttons = pavian:::common_buttons(set_name))) %>%
      pavian:::formatSummaryDT(display_percent = TRUE)),
  tabPanel("Raw read numbers",
    DT::datatable(samples_summary, rownames = FALSE,
                  class = "stripe hover row-border compact",
                  extensions = 'Buttons',
                  options = list(dom = 'Bfrtip',
                                 buttons = pavian:::common_buttons(set_name))) %>%
      pavian:::formatSummaryDT(display_percent = FALSE)),
  tabPanel("Sample information",
    DT::datatable(sample_data,
                  class = "stripe hover row-border compact nowrap",
                  extensions = 'Buttons',
                  options = list(dom = 'Bfrtip',
                                 buttons = pavian:::common_buttons(set_name))))
)
```

## Read-fate overview

```{r qc_plot, fig.width=9, fig.height=4.5}
tryCatch({
  library(ggplot2)
  cols_present <- colnames(samples_summary)
  pick <- function(re) cols_present[grep(re, cols_present, ignore.case = TRUE)][1]

  qc <- data.frame(
    Sample       = rownames(samples_summary),
    Unclassified = samples_summary[, pick("unclassified")],
    Chordate     = samples_summary[, pick("chordate")],
    Microbial    = samples_summary[, pick("microbial")],
    stringsAsFactors = FALSE
  )
  qc_long <- reshape2::melt(qc, id.vars = "Sample",
                            variable.name = "Category", value.name = "Reads")
  qc_long$Category <- factor(qc_long$Category,
                             levels = c("Unclassified","Chordate","Microbial"))
  ggplot(qc_long, aes(Sample, Reads, fill = Category)) +
    geom_col(position = "fill") +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_fill_manual(values = c("Unclassified" = "#b0b0b0",
                                 "Chordate"     = "#f4a261",
                                 "Microbial"    = "#2a9d8f")) +
    labs(y = "Fraction of reads", x = NULL, fill = NULL,
         title = "Read fate per sample") +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title  = element_text(face = "bold"))
}, error = function(e) {
  cat("*(Could not render QC plot:", conditionMessage(e), ")*")
})
```

Classification results
=====================================  

```{r classifications}
merged_reports <- pavian::merge_reports2(reports, col_names = sample_data$Name)
taxonReads <- merged_reports$taxonReads
cladeReads <- merged_reports$cladeReads
tax_data   <- merged_reports[["tax_data"]]

sel_bacteria <- grepl("d_Bacteria",  tax_data[,"taxLineage"]) & tax_data[,"taxRank"] == 'S'
sel_viruses  <- grepl("d_Viruses",   tax_data[,"taxLineage"]) & tax_data[,"taxRank"] == 'S'
sel_fungi    <- grepl("k_Fungi",     tax_data[,"taxLineage"]) & tax_data[,"taxRank"] == 'S'
sel_euk      <- grepl("d_Eukaryota", tax_data[,"taxLineage"]) & tax_data[,"taxRank"] == 'S'
sel_protists <- grepl("d_Eukaryota", tax_data[,"taxLineage"]) & !sel_fungi &
                !grepl("p_Chordata", tax_data[,"taxLineage"]) & tax_data[,"taxRank"] == 'S'

my_df <- data.frame(
  Name    = tax_data$name,
  Max     = apply(cladeReads, 1, max, na.rm = TRUE),
  cladeReads,
  Lineage = pavian:::beautify_taxLineage(tax_data$taxLineage),
  stringsAsFactors = FALSE
)
my_dt <- function(sel) {
  selw <- which(sel)
  max_order <- order(my_df[sel, "Max"], decreasing = TRUE)
  if (sum(sel) > 100) {
    selw <- selw[max_order[1:100]]
    h <- sprintf("Showing top 100 of %s species.", sum(sel))
  } else {
    selw <- selw[max_order]; h <- ""
  }
  shiny::tagList(
    shiny::HTML(h),
    DT::datatable(my_df[selw,, drop = FALSE], height = "600px", rownames = FALSE,
                  class = "stripe hover row-border compact nowrap",
                  extensions = 'Buttons',
                  options = list(dom = 'Bfrtip',
                                 buttons = pavian:::common_buttons(set_name),
                                 scrollX = TRUE))
  )
}

tabsetPanel(
  tabPanel("Bacteria",            my_dt(sel_bacteria)),
  tabPanel("Viruses",             my_dt(sel_viruses)),
  tabPanel("Eukaryotes",          my_dt(sel_euk)),
  tabPanel("Eukaryotes / Fungi",  my_dt(sel_fungi)),
  tabPanel("Eukaryotes / Protists", my_dt(sel_protists))
)
```

## Top hits per sample

Top 10 species with ≥ 50 clade reads, after removing Chordata and synthetic constructs.

```{r top_hits}
top_n     <- 10
min_reads <- 50

do.call(tabsetPanel, lapply(names(reports), function(n) {
  r <- reports[[n]]
  r <- r[r$taxRank == "S" & r$cladeReads >= min_reads, , drop = FALSE]
  r <- r[!grepl("Chordata|synthetic", r$taxLineage), , drop = FALSE]
  r <- r[order(-r$cladeReads), , drop = FALSE]
  r <- utils::head(r, top_n)
  tabPanel(n,
    if (nrow(r) == 0)
      HTML("<i>No species above threshold.</i>")
    else
      DT::datatable(
        data.frame(
          Species    = sub("^._", "", r$name),
          CladeReads = r$cladeReads,
          TaxonReads = r$taxonReads,
          Lineage    = pavian:::beautify_taxLineage(r$taxLineage),
          stringsAsFactors = FALSE
        ),
        rownames = FALSE,
        options  = list(dom = "t", pageLength = top_n)
      )
  )
}))
```

## Cross-sample heatmap

Top 40 microbial species across all samples (log10 of clade reads).

```{r heatmap, fig.width=10, fig.height=8}
tryCatch({
  sel_microbe <- sel_bacteria | sel_viruses | sel_fungi
  mat <- as.matrix(cladeReads[sel_microbe, , drop = FALSE])
  rownames(mat) <- sub("^._", "", tax_data$name[sel_microbe])

  # Replace NA/NaN/Inf with 0
  mat[!is.finite(mat)] <- 0

  # Drop rows/cols that are entirely zero (clustering can't handle them)
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  mat <- mat[, colSums(mat) > 0, drop = FALSE]

  # Keep top 40 by max abundance
  mat <- mat[order(-apply(mat, 1, max)), , drop = FALSE]
  mat <- mat[seq_len(min(40, nrow(mat))), , drop = FALSE]

  # Need at least 2 rows AND 2 cols for clustering
  do_cluster_rows <- nrow(mat) >= 2
  do_cluster_cols <- ncol(mat) >= 2

  if (nrow(mat) < 1 || ncol(mat) < 1) {
    cat("*(No non-zero microbial abundance to plot.)*")
  } else {
    pheatmap::pheatmap(
      log10(mat + 1),
      color        = colorRampPalette(c("white","#a8c8ec","#1f4e79"))(50),
      cluster_rows = do_cluster_rows,
      cluster_cols = do_cluster_cols,
      fontsize_row = 8,
      fontsize_col = 9,
      main         = "log10(clade reads + 1) — top microbes"
    )
  }
}, error = function(e) {
  cat("*(Heatmap not rendered:", conditionMessage(e), ")*")
})
```

## Prevalence-based contaminant flagging

Species detected in *all* samples at low level may be reagent/lab contaminants; species exclusive to a single sample are more likely real.

```{r contam_flag}
tryCatch({
  sel_microbe <- sel_bacteria | sel_viruses | sel_fungi
  m <- as.matrix(cladeReads[sel_microbe, , drop = FALSE])
  rownames(m) <- sub("^._", "", tax_data$name[sel_microbe])
  prevalence <- rowSums(m > 10, na.rm = TRUE)
  max_reads  <- apply(m, 1, max, na.rm = TRUE)
  flagged <- data.frame(
    Species    = rownames(m),
    Prevalence = prevalence,
    MaxReads   = max_reads,
    Pattern = ifelse(
      prevalence == ncol(m) & max_reads < 1000, "Likely contaminant",
      ifelse(prevalence == 1 & max_reads >= 100, "Sample-specific", "")
    ),
    stringsAsFactors = FALSE
  )
  flagged <- flagged[flagged$Pattern != "", , drop = FALSE]
  flagged <- flagged[order(flagged$Pattern, -flagged$MaxReads), , drop = FALSE]
  DT::datatable(flagged, rownames = FALSE,
                options = list(pageLength = 20, dom = 'Bfrtip',
                               buttons = pavian:::common_buttons(set_name)),
                extensions = 'Buttons')
}, error = function(e) {
  cat("*(Contaminant flagging skipped:", conditionMessage(e), ")*")
})
```

Sankey visualization
=====================================  

```{r sankey_vis}
if (isTRUE(params$include_sankey)) {

  tax_taxRanks <- c("D","K","P","C","O","F","G","S")

  all_names <- sub("^._", "", sort(unique(unlist(
    sapply(reports, function(x) x$name[x$taxRank != "-"])
  ))))
  # Sanitize for JS string injection
  all_names <- gsub('"', "'", all_names, fixed = TRUE)
  all_names <- gsub("\\\\", "/", all_names)

  # Explicit 20-color palette (category20b colors) — independent of D3 version
  palette <- c(
    "#393b79","#5254a3","#6b6ecf","#9c9ede",
    "#637939","#8ca252","#b5cf6b","#cedb9c",
    "#8c6d31","#bd9e39","#e7ba52","#e7cb94",
    "#843c39","#ad494a","#d6616b","#e7969c",
    "#7b4173","#a55194","#ce6dbd","#de9ed6"
  )
  colourScale <- sankeyD3::JS(sprintf(
    "d3.scaleOrdinal().range([%s]).domain([%s])",
    paste0('"', palette,                 '"', collapse = ","),
    paste0('"', c(all_names, "other"),   '"', collapse = ",")
  ))

  if (length(params$filter_tax) != 0 && isTRUE(is.na(params$filter_tax))) {
    filt <- c("Chordata", "other sequences")
  } else {
    filt <- params$filter_taxa
  }

  # Sankey navigation TOC
  toc_html <- paste0(
    "<p><b>Jump to:</b> ",
    paste(sprintf("<a href='#sankey-%s'>%s</a>",
                  gsub("[^A-Za-z0-9_-]", "_", names(reports)),
                  names(reports)),
          collapse = " &middot; "),
    "</p>"
  )

  htmltools::tagList(
    HTML(toc_html),

    HTML(
      "<div style='border:1px solid #ddd; padding:8px 12px; background:#fafafa;
                  font-size:90%; margin-bottom:1em; border-radius:4px;'>
         <b>How to read the Sankey:</b> node width is proportional to clade reads;
         node color is assigned consistently per taxon across samples;
         link color follows the source node. Filtered taxa appear in the run
         information section above.
       </div>"
    ),

    # Shim for missing Shiny + color scale + post-render colorizer
    HTML(sprintf(
      "<script>
         if (typeof Shiny === 'undefined') {
           window.Shiny = { onInputChange:function(){}, setInputValue:function(){},
                            addCustomMessageHandler:function(){}, bindAll:function(){},
                            unbindAll:function(){} };
         }
         sankey_colorscale = %s;

         function pavianRecolor() {
           if (typeof sankey_colorscale !== 'function') return;
           document.querySelectorAll('.node').forEach(function(node) {
             var rect = node.querySelector('rect');
             var text = node.querySelector('text');
             if (!rect || !text) return;
             var name = text.textContent.trim();
             var c = sankey_colorscale(name);
             rect.setAttribute('fill', c);
             rect.style.fill = c;
           });
           document.querySelectorAll('.link').forEach(function(link) {
             var cls = (link.getAttribute('class') || '')
                         .split(/\\s+/).filter(function(x){ return x && x !== 'link'; });
             if (cls.length) {
               var c = sankey_colorscale(cls[0]);
               link.style.stroke = c;
               link.style.strokeOpacity = 0.4;
             }
           });
         }
         window.addEventListener('load', function(){ setTimeout(pavianRecolor, 300); });
         setTimeout(pavianRecolor, 1500);
         setTimeout(pavianRecolor, 3000);
       </script>",
      colourScale
    )),

    lapply(names(reports), function(n) {
      my_report <- reports[[n]]
      for (f in filt) my_report <- filter_taxon(my_report, f)
      anchor <- gsub("[^A-Za-z0-9_-]", "_", n)
      htmltools::tagList(
        HTML(sprintf("<h2 id='sankey-%s'>%s</h2>", anchor, n)),
        pavian:::build_sankey_network(
          my_report,
          nodePadding     = 13,
          xScalingFactor  = .9,
          nodeStrokeWidth = 0,
          zoom            = FALSE,
          colourScale     = "sankey_colorscale",
          LinkGroup       = "source_name"
        )
      )
    })
  )
}
```

About
=====

This file was generated with the Pavian R package version `r utils::packageVersion('pavian')` on `r date()`.
Please cite Pavian if you use it in your research.

<details><summary>Software environment</summary>
<pre>
```{r session, echo=FALSE, results='asis'}
cat(paste(capture.output(sessionInfo()), collapse = "\n"))
```
</pre>
</details>
